---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  eval = TRUE
)
```

# rexpanse1

<!-- badges: start -->
<!-- badges: end -->

The goal of rexpanse1 is to provide the necessary functions for generating the results of the paper *Socioeconomic inequalities in urban exposome profiles across the life span in European cohorts*.

## Installation

You can install the latest version of rexpanse1 like so:

``` r
# install.packages("remotes")
remotes::install_github("expanse-wp-3/rexpanse1")
```

## Getting started

### Data

For illustration, we will load the example data set, `expanse`, included with the package.

```{r data}
library(rexpanse1)

# Example data set
data(expanse)
str(expanse)
```

For producing results for your cohort you should use your data instead.

>**Important**: Note that exposure variables must be named as in the example data set and as shown below.

| Variable name    | Original WP1 name   | Description                                          | Year | Version |
|------------------|---------------------|------------------------------------------------------|------|---------|
| `no2`            | `NO22010`           | Annual average NO2 concentration                     | 2010 | 1       |
| `pm25`           | `PM252010`          | Annual average PM2.5 concentration                   | 2010 | 1       |
| `bc`             | `BC2010`            | Annual average BC concentration                      | 2010 | 1       |
| `o3`             | `O32010w`           | Warm season average O3 concentration                 | 2010 | 1       |
| `ndvi`           | `ndvi_2019`         | NDVI                                                 | 2019 | 1       |
| `imperv`         | `ImpSurf_2015`      | Percentage of impervious surfaces                    | 2015 | 1       |
| `dist_water`     | `Dist_water`        | Distance to water (km)                               |      | 1       |
| `temp_cold_mean` | `temp_2010_cold`    | Mean temperature during warm season (May-October)    | 2010 | 1       |
| `temp_cold_sd`   | `temp_2010_cold_sd` | Temperature standard deviation during warm season    | 2010 | 1       |
| `temp_warm_mean` | `temp_2010_warm`    | Mean temperature during cold season (November-April) | 2010 | 1       |
| `temp_warm_sd`   | `temp_2010_warm_sd` | Temperature standard deviation during cold season    | 2010 | 1       |


### Results

All the results can be produced by running a single function, `run_pca()`, with the appropriate arguments. See `?run_pca` for details.

Using the example data set:

```{r}
# I create a temporary directory to store the results.
# You do not need to do this.
my_results_dir <- tempdir()

# Run entire analysis.
catalonia_pca <- run_pca(
  expanse,
  ses_var = "ses_cat",
  cohort_name = "catalonia",
  write = TRUE,
  output_dir = my_results_dir
)
```

Note that you may get some warnings informing you that some of the principal components have been multiplied by -1 to make the interpretation somewhat consistent across the cohorts.

Principal components are transformed this way if the loading for `no2` or `ndvi` are negative.

You should pass your data frame formatted as decribed in the previous section.

Note that when `write = TRUE`, you need to specify an appropriate directory (folder) path to `output_dir`. The defult directory is the current working directory.

### Details

`run_pca()` returns a list with the following elements.

#### Cohort name

The cohort name specified:

```{r}
catalonia_pca$cohort
```

#### Exposure quantiles

A set of quantiles of the exposure variables:

```{r}
catalonia_pca$quantiles
```

#### PCA by domain summary

A summary of the PCA by domain in long format for easy plotting.

```{r}
head(catalonia_pca$pca_summary)
```

#### Associations with SES

The estimated coefficients for the linear models with each principal component as outcome and the SES variable supplied to `ses_var`. There is an option to include more variables using the `other_vars` argument in `run_pca()`.

```{r}
head(catalonia_pca$pc_ses_models)
```

#### Package version

The version of the rexpanse1 package that was used to produce the results.

```{r}
catalonia_pca$package_version
```
