---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# rexpanse1

<!-- badges: start -->
<!-- badges: end -->

The goal of rexpanse1 is to provide the necessary functions for generating the results of the paper *Socioeconomic inequalities in urban exposome profiles across the life span in European cohorts*.

## Installation

You can install the latest version of rexpanse1 like so:

``` r
# install.packages("remotes")
remotes::install_github("expanse-wp-3/rexpanse1")
```

## Getting started

For illustration, we will load the example data set, `exposures`, included with the package.

```{r data}
library(rexpanse1)

# Example data set
data(exposures)
str(exposures)
```

You should use your data instead. Note that exposure variables must be named as in the example data set:

| Variable name    | Original name | Domain            | Description                          | Year | Version |
|------------------|---------------|-------------------|--------------------------------------|------|---------|
| `no2`            | `NO2FULLt`    | Air pollution     | Annual average NO2 concentration     | 2010 | Pilot   |
| `pm25`           |               | Air pollution     | Annual average PM2.5 concentration   | 2010 | Pilot   |
| `bc`             |               | Air pollution     | Annual average BC concentration      | 2010 | Pilot   |
| `o3`             |               | Air pollution     | Warm season average O3 concentration | 2010 | Pilot   |
| `ndvi`           |               | Built environment | NDVI                                 | 2019 | Pilot   |
| `imperv`         |               | Built environment | Percentage of impervious surfaces    |      | Pilot   |
| `dist_water`     |               | Built environment | Distance to water (km)               |      | Pilot   |
| `temp_cold_mean` |               |                   |                                      |      |         |
| `temp_cold_sd`   |               |                   |                                      |      |         |
| `temp_warm_mean` |               |                   |                                      |      |         |
| `temp_warm_sd`   |               |                   |                                      |      |         |

### Quantiles

Get quantiles necessary for producing boxplots:

```{r}
expo_quantiles <- get_quantiles(data = exposures, write = FALSE)
```

### PCA

Get a *summary* (PC loadings and variance explained) of the principal component analysis by domain:

```{r}
pca_summary <- get_pca(data = exposures, summary = TRUE)
str(pca_summary)
```

Get the PCA summary as above, as well as the actual principal component values attached to the original exposures data frame.

```{r}
pca_full <- get_pca(data = exposures, summary = FALSE)
str(pca_full)
```

### Associations with SES variable

We can obtain the coeficient estimates from a linear regression with a principal component as outcome and the categorical SES variable as predictor with:

```{r}
ses_fit <- fit_pc_model(
  data = pca_full$data_merged,
  pc_var = "pc1_air",
  ses_var = "ses_cat"
)
```
